---
title: "Getting started with rswat"
author: Dean Koch
date: July 27, 2023
output: 
  github_document:
    toc: TRUE
bibliography: rswat_intro.bib
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This article explains how to use rswat to explore and manage a SWAT+ project. It picks up where the [Lamar River tutorial](https://github.com/deankoch/rswat.maker/blob/master/vignettes/articles/lamar.md) from [`rswat.maker`](https://github.com/deankoch/rswat.maker) left off, by showing how to edit parameters and carry out simulations. Our example is the small headwater catchment of Soda Butte Creek, near the Silver Gate boundary in the northeast corner of Yellowstone National Park.

```{r setup, include=FALSE}
library(rswat)
library(rswat.maker)
# rmarkdown::render('D:/rswat/vignettes/articles/rswat_intro.Rmd')

# wipe existing output directory
dest_dir = 'D:/rswat_data/swat/soda_test'
if( dir.exists(dest_dir) ) unlink(dest_dir, recursive=TRUE)
dir.create(dest_dir, recursive=TRUE)
```

```{r source-project, cache=TRUE, echo=FALSE, fig.align='center', fig.dim=c(5, 5)}
# load the headwater subcatchment TxtInOut (SWAT+ project) from the Lamar vignette
lamar_dir = 'D:/rswat_data/lamar'
soda_dir = save_split(lamar_dir)[['sub']][2]
soda_qswat = run_qswat(soda_dir)[['output']] |> readLines() |> jsonlite::fromJSON()

# plot basin and overlay sub-basins found with SWAT 
sub_dir = plot_qswat(soda_dir, quiet=TRUE, main='Soda Butte headwater (QSWAT+)')

# get paths required below
txt_dir = soda_qswat[['txt']]
swat_path = soda_qswat[['simulator_dir']] |>
  file.path('SWATPlusEditor/resources/app.asar.unpacked/static/swat_exe/rev60.5.7_64rel.exe')
```

## Getting started

Users will need a pre-existing SWAT+ project directory. This is a collection of dozens to hundreds of plaintext config files, usually in a directory called "TxtInOut" created by QSWAT+ and SWAT+ Editor. Check out [`rswat.maker`](https://github.com/deankoch/rswat.maker) for an R-based workflow to run these programs in batch mode.

It can be hard to track changes with so many files so we recommend always working with backups, and include helper functions `rswat_backup` and `rswat_restore` to make this easier. For the Soda Butte example we will start by making a copy of our project directory (with path `txt_dir`) in a temporary location (`dest_dir`) by passing their paths to `rswat_restore` 

```{r source-copy}
# the source and destination directory names
basename(txt_dir)
basename(dest_dir)

# make the copy
txt_dir |> rswat_restore(dest_dir, overwrite=TRUE) |> head()
```

## Open a project

Open any SWAT+ project by passing its path to `rswat`. The code below loads the copy we just created 

```{r intro-initialize-1}
rswat(swat_dir=dest_dir, quiet=TRUE)
```
Here we have set `quiet` to disable the progress bar, which doesn't render well in markdown. Call `rswat` again without arguments to get a summary of the project

```{r intro-initialize-2}
rswat()
```

The "all requested files loaded" here refers to optional arguments `include` and `exclude` whose default settings skip loading certain large files, like weather inputs, until they are needed. Read more about this in `?rswat`. 

# Exploring SWAT+

For a process-based model, SWAT+ has an unusually large number of parameters (The latest input documentation PDF runs over 250 pages in length!). This can make for a steep learning curve. rswat provides tools to help users to organize this information and get oriented with their config files.

## File index

Get a list of all files in your SWAT+ project, loaded or not, using `rswat_files`

```{r intro-files-1}
rswat_files()
```

This can be filtered using a number of different criteria (read more about this in `?rswat_files`)  

```{r intro-files-2}
rswat_files('climate')
```

## Open config files

To open a file, pass the file name to `rswat_open`. For example, the 'time.sim' file controls the time period of simulations, and consists of a single, one-line table. `rswat_open` returns it as a single data frame

```{r open-time}
rswat_open('time.sim') |> str()
```

Most SWAT+ config files have this very simple single-table structure. The few multi-table exceptions, such as `print.prt`, are returned as a list of data frames. For example this code prints the first two tables

```{r open-print}
rswat_open('print.prt') |> head(2)
```

## Open output files

Outputs are tabular so they can be browsed the same way as config files, using `rswat_files` and `rswat_open`. For example, the yearly water balance file in our example looks like this

```{r simulate-result-pre}
rswat_open('basin_wb_yr.txt') |> rswat_date_conversion() |> dplyr::tibble()
```

`dplyr::tibble` is useful for printing these results in R, as the output tables are often very long (for daily outputs at least) and too wide to print without wrapping. `rswat_date_conversion` translates the Julian date columns to an R Date column. 

## Search

Once a file has been loaded, it becomes searchable with `rswat_find`. This supports fuzzy matching to variable names, to help with typos and names that have changed over time. 

For example the potential evapotranspiration (PET) model flag, formally called 'IPET' in SWAT2012, is now called 'pet'. Searching for the old name turns up no exact or substring matches (the default search mode). However, with `fuzzy=2` we find the new name, plus one false positive

```{r intro-find}
rswat_find(pattern='IPET', fuzzy=2)
```

Repeating the call with `fuzzy=1` produces the unique result 'pet'. 

We can see that this variable is stored in the file 'codes.bsn'. Pass a file name in `pattern` to report on all the variables in the file

```{r intro-find-file}
rswat_find(pattern='codes.bsn')
```
 
By default `rswat_find` will attempt to match all results to definitions (the `desc` field) from the SWAT+ inputs documentation PDF. As there is some uncertainty in matching old and new names automatically to text scraped from a PDF, a three-star match confidence ranking is reported by `rswat_find` along with any aliases. When `alias` is an empty string it means the function has found an exact match, and the definition is very likely to be correct.

## Documentation

rswat includes a plaintext copy of latest SWAT+ inputs documentation PDF ("inputs_swatplus_rev60_5.pdf", last downloaded from the SWAT+ website in October, 2022) and makes it searchable via `rswat_docs`. Call this function to get a `tibble` of results for a search query.

```{r intro-docs-1}
rswat_docs('codes.bsn')
```

Searching for a file name will usually pull up a full definitions list, along with some information about where in the PDF this text can be found. Documentation can also be searched with `rswat_docs`. For example the pattern 'IPET' turns up a new match: 'ipet' in the 'hru-lte.hru' file. 

```{r intro-docs-2}
rswat_docs('IPET')
```

Our example is not an "LTE"-type project, so we don't have 'hru-lte.hru' in our project directory. `rswat_docs` will find the match but `rswat_find`, which searches only loaded files, will not.

Those who are new to SWAT+ or have trouble remembering its many variable names can also search definition text for matches to keywords by setting `defs=TRUE`. For example, the search pattern 'evapotranspiration' turns up both of the IPET-type parameters along with a few other results for related processes
```{r intro-docs-3}
rswat_docs('evapotranspiration', defs=TRUE)
```


# Editing

So far we have only browsed the files for an existing project. The real power of rswat is in allowing R users to easily modify the parameters in the SWAT+ config files on disk. rswat makes this more intuitive for R users by representing all config files as data frames 

## Modify parameters

The 'print.prt' file is a good place to start editing as it controls which time period to print to the output files. The result will be immediately obvious the next time you run a simulation.

In the earlier `rswat()` summary, and in the existing output file above, we can see that the print period is currently set to a single date. This is because 'print.prt' is currently configured for a one year long burn-in period during which the output is muted.

```{r nyskip-docs}
rswat_find('nyskip')
```

To modify a parameter, get a copy of its data frame with `rswat_open`, change its value, and pass the data frame back to `rswat_write`. The code below shows how to remove this burn-in period via its parameter `nyskip`

```{r modify-print-1}
# copy the relevant table
prt = rswat_open('print.prt')[[1]]

# assign new values
prt[['nyskip']] = 0L
prt |> rswat_write(overwrite=TRUE)
```

Calling `rswat_write` with default `overwrite=FALSE` will report the pending changes but not write them

## Project state

Changes written with `rswat_write` persist in the files on disk, and subsequent calls to `rswat_open` will return the newest version. `rswat()` calls will also reflect the current state on disk

```{r modify-print-result}
rswat()
```
Notice the summary shows the updated print range for 'print.prt' (now matching 'time.sim').

rswat uses a reference class (R5) internally to keep an up-to-date database of information on the loaded SWAT+ project. This is important as it allows all rswat function to have a common knowledge about the state of the files on disk without passing big lists around in function calls, and it allows us to more easily implement a cache to speed up text parsing/writing.

The data frames returned by `rswat_open`, however, are ordinary R data frames with copy-on-modify semantics. This means rswat will ignores modifications to the data frames until you pass them to `rswat_write`, to commit the changes to disk.

# Simulations

For running simulations, rswat uses `shell` to call the simulator executable. Users will need to supply their local path to this (.exe) file by passing it as argument `exe_path` in an `rswat` call.

```{r exe-path}
rswat(exe_path=swat_path)
```

In the latest release the simulator file for Windows is called 'rev60.5.7_64rel.exe' and it can be found in the 'SWAT/SWATPlus/SWATPlusEditor' directory tree. Standalone versions can also be downloaded.

Note that most of the package functions will work without setting this path, which means rswat still functions as an editor without a working SWAT+ installation. Once the path is set, users can call `rswat_exec()` to run the simulator and write output to the project directory.

```{r simulate}
rswat_exec()
```

The function returns a list of output files written when it finishes. If we open the yearly water balance file again, we can see that the results from 2022 (formerly a hidden burn-in year) now appear in the output, as a result of our earlier change to 'print.prt'.

```{r simulate-result-post}
rswat_open('basin_wb_yr.txt') |> rswat_date_conversion() |> dplyr::tibble()
```

# Applications

`rswat_open`, `rswat_write`, and `rswat_exec` is a powerful combination. It allows users to programmatically control virtually all aspects of a SWAT+ simulation. In the next vignette we give a few examples, showing how to calibrate a forecasting model for the Soda Butte catchment.
