
Dean Koch
February 06, 2023

This document introduces the `rswat` R package, showing off the important existing features and functionality. Maybe this can serve as a draft for a paper later on.

## Notes

This package is for SWAT+ users but it is not affiliated with the SWAT+ development team. It is an independent project by an R user interested in SWAT+. Don't bother the SWAT+ developers with `rswat` questions, and vice-versa.

`rswat` is designed for SWAT+ (read “swat-plus”), and not its predecessor “SWAT2012”. Both versions implement essentially the same model, and the name "SWAT" is often used interchangeably for both the model and (both of) the simulators. However the older simulator uses a different configuration file scheme that has not been tested with `rswat`. 

## rswat Features

* summarize/load SWAT+ project files using `rswat()`
* back-up and restore projects with `rswat_backup()`
* execute simulations with `swat_exec()`
* open simulation outputs as data frames with `rswat_open()`
* fetch any parameter table as a data frame with `rswat_open()`, save changes with `rswat_write()`
* fuzzy search for files/parameters with `rswat_find`
* fuzzy search for names/descriptions in documentation PDF with `rswat_docs`
* script-friendly helpers for parameter tuning/fitting 


## What is SWAT+?

SWAT+ is a program for generating detailed accounts of water movement through a watershed by means of a model simulation. SWAT+ builds a digital representation of the initial state of a watershed, then steps the model forward in time, calculating future states based on the present state and environmental drivers (parameters) of the user's choosing. Hydrologists analyze SWAT+ outputs under different parameterizations to study the dynamics of water channels, reservoirs, and their associated catchments.

## Why rswat?

Watersheds are sprawling, intricate systems, with many different components that interact with each other. As a result, there is a lot of quantitative information to manage and understand when working with SWAT+. `rswat` is designed to help users manage all these data, and to enable the automation of model configuration and prediction using reproducible R scripts.


## SWAT+ sub-processes

Most of the water entering a watershed arrives as precipitation, which drains along one of many paths through the landscape towards an exit point, such as on a creek or river. Building a detailed water budget is essentially a matter of monitoring how much water arrives over time (eg. with weather stations), and how much leaves (eg. with stream gauges), then deducing what has happened in between using principles from hydrology.

This "in between" involves not just a stream network, but also a host of landscape features that play a significant role in storing and moving water, including snow pack, soils, aquifers, ponds and lakes. These components interact with each other, and with the surrounding environment, so an accurate model has to account for transfers between them, as well as losses from evaporation and evapo-transpiration, and leakage into deep aquifers

SWAT+ will simulate all of these sub-processes (and more), making the appropriate adjustments to the water budget so that downstream values remain accurate. However each adjustment requires some thoughtful configuration to reflect the unique properties of different watersheds.


## Inputs and outputs

In simulations, each SWAT+ sub-process is defined by one or more configuration files. These files must be filled with parameter values, and placed somewhere they can be found by the simulator executable. SWAT+ writes its output files (in plain text) to the same location, so the directory is usually named "TxtInOut". I will refer to it as the "project directory".

The code below species the project directory for an example we will look at later

```{r project directory}
swat_dir = 'D:/rswat/example_projects/blacktail'
swat_files = list.files(swat_dir)
```

there are quite a few files (all plain text)

```{r project files count}
length(swat_files)
```

Here are some of the file extensions you will likely encounter when working with SWAT+

```{r project directory extensions}
swat_files |> tools::file_ext() |> unique()
```

A typical SWAT+ model can have thousands of model parameters spread over several dozen configuration files in the project directory. A potentially large number of weather inputs (with extensions .hmd, .sol, .pcp, .tmp, .wnd) and model outputs (.txt) add yet more files to the mix.

`rswat` aims to simplify the management of this data, and to assist with parameter estimation and tuning, all from within R.


## Simulations

`rswat` is in essence a text file editor/manager. It does no simulations of its own. Instead it calls the SWAT+ simulator (.exe) file to run simulations and write output files. This binary must be installed separately by following the instructions linked on the SWAT+ homepage.

Below, I've set the path for the installation on my local machine then called the simulator manually. 

```{r set exe path}
exe_path = 'D:/rswat/swat_executable/rev60.5.4_64rel.exe'
shell(exe_path, intern=TRUE)
```

This prints version information, but (being in the wrong working directory) fails to locate the configuration files needed to run the simulation. In examples later on, we will use `swat_exec` to call the simulator the right way. 

## Dependencies

We have tried to keep `rswat` light on dependencies. There is no reliance on SWAT+ Editor or QSWAT+, though we recommend them for your own use. The most important R dependencies are `data.table` (to load text files) and `dplyr` (to help with parsing). 

```{r deps, message=FALSE}
library(data.table)
library(dplyr)
# library(rswat)

# TEMPORARY: load development version of rswat
library(devtools)
document()

```

**You will also need an existing SWAT+ project to work on.** Later on we may add a function for building a SWAT+ model from scratch, similar to the SWAT+AW project. For now we recommend using the QSWAT+ GUI to initialize a copy of your SWAT+ project files.

Pass the directory path (usually named "TxtInOut") to `rswat` to assign your project of interest. By default this call loads some of the more important config files, and it produces a bunch of console output about its progress (hidden in this document for tidiness). 

```{r set project directory, results='hide', message=FALSE}
rswat(swat_dir)
```

Pass a new project directory path at any time to switch to a different project, or a previous version. You can also set/change `exe_path` at any time to enable simulations with a specific version of the SWAT+ simulator.

```{r execution setup, results='hide', message=FALSE}
rswat(exe_path=exe_path)
```


## Summarize a SWAT+ project

Once the project directory is assigned, subsequent calls to `rswat` without arguments will produce a short summary of the currently loaded project. 

```{r list project directory, message=FALSE}
rswat()
```

This reports the following:

- **directory** the currently loaded project directory
- **simulator** the executable to use for simulations
- **climate** the number of weather stations, and the name of a file listing weather inputs
  - **wgn** which (if any) weather variables will be simulated
  - **sta** which (if any) weather variables will be loaded from input files
- **time.sim** the time period over which the simulation will run, including burn-in time
- **print.prt** the time period over which simulation outputs will be written to output files
  - the output files to be written and their file names (if any) 
- **file.cio** the number of configuration files detected and the number of categories

Let's unpack this information in reverse order:

#### file.cio

The most important file in a SWAT+ project is named **file.cio**. It lists - or points to other files that list - every configuration file to be loaded by the simulator. It also groups important files into categories.


#### Time periods

Because most SWAT+ state variables are initialized with a guess, simulations tend to improve following a burn-in period. This is an initial period during which the model's internal state variables evolve to more realistic values, reflecting, among other things, normal seasonal trends and the echoes of past weather events. SWAT+ therefore provides two time period settings, one in **time.sim** - the period to simulate - and one in **print.prt** - the period to produce outputs, a subset of the simulation period.
  
SWAT+ is written in Fortran, which is fast, but all of the input/output happens via plain text files, which can be large and slow to write. As a result you may want to disable unnecessary output files, particularly if your basin has many HRUs, or when simulating long periods. In this example we have set up the **print.prt** file to request only a single output file ("channel_sd.txt", daily values). The other outputs still exist in the simulation, but you won't see them because we have requested they not be printed. 

#### Weather

The **climate** entry reports on weather stations, **sta**, and generators, **wgn**. Ideally you should supply daily weather data to drive your simulation. SWAT+ accepts regional weather inputs at the level of the HRU through point locations called "weather stations" (they don't have to correspond to real stations). Each HRU is mapped to exactly one of these stations - the closest one - and each mapped station should have one or more accompanying weather records.
  
SWAT+ has a built-in system for generating missing weather data. This system gets activated automatically for any variable that is missing a weather input file. In our example "slr" is missing, which means SWAT+ will generate daily solar radiation values randomly (using monthly climatic means and variances).

The creation of regionalized weather data is an interesting and tricky statistical problem. Readers may be interested in my other R package snapKrig for an interpolation tool based on Gaussian processes. We will come around to this topic later in a detailed example. 

## Listing files

To browse files in your project directory by name and type, check out `?rswat_files`. When called without arguments, this returns a tibble describing all (recognized) files belonging to a SWAT+ project. This includes configuration files, *gwflow* and weather inputs, outputs, logs, and `rswat` backups.

```{r list all files and types, message=FALSE}
rswat_files() |> pull('type') |> unique()
```


## Backing up or restoring a project

It is all too easy to violate syntax and parameter range specifications in SWAT+, and the result is usually some inscrutable Fortran error message and a failure to produce output. Don't let this dissuade you from experimenting with the model configuration files. Just use backups.

A copy of the essential configuration files will amount to very little disk space usage in most cases, and `rswat` makes it easy to copy your project at any time with `rswat_backup()`. 

Before we start tinkering with our project, we will make a backup zip file for recovering all of the configuration files (and you should too!)

```{r backup creation}
rswat_backup(zip=T, include='config')
```

Weather input files, which can be very large, are not included in the "config" group. You can also omit the `include` argument to zip all recognized files in the project directory (including all weather and output files). See `?rswat_backup` and `?rswat_files` for more information on the `include` and `exclude` arguments.

```{r backup creation}
rswat_files(include='weather')
```


## Opening and editing parameter tables

Say we are happy with the SWAT+ project as it is configured, except that we want to simulate a shorter time period. We will need to make changes to the two timing configuration files, "time.sim" and "print.prt". We start by opening them to view the contents.

`rswat_open` will open any SWAT+ file, returning a copy of the parameter data inside as an R data frame (or a list of them). Here we open the "time.sim" file, which controls the time period of the simulation.

```{r open time file}
time.sim = rswat_open('time.sim')
time.sim
```

The R data frame need not be assigned the same name as the file, like we did here, though it seems like the most obvious name to use.

Some configuration files contain multiple table, so they are returned as a list of data frames. "print.prt" is an important example:

```{r open print file}
print.prt = rswat_open('print.prt')
class(print.prt)
length(print.prt)
```

We are interested in the first table from "print.prt", which controls the time period for printing outputs

```{r extract print table}
print.prt.t1 = print.prt[[1]]
print.prt.t1
```


The simulation is currently set up to run the period 1988-1993 and omit the first year from output (burn-in). We will instead request to simulate only the first year, and return all of it. Thus we modify the pair of date fields in each table, as well as the "nyskip" (burn-in) field from "print.prt".

```{r change timing}
# change simulation period
time.sim[['yrc_end']] = 1989
time.sim[['day_end']] = time.sim[['day_start']]

# change printout period
print.prt.t1[['nyskip']] = 0
print.prt.t1[['yrc_end']] = 1989
print.prt.t1[['day_end']] = print.prt.t1[['day_start']]
```

## Save changes

To write changes to disk, pass the modified data frame back to `rswat_write`. `rswat` will report the corresponding modifications to the configuration file using the correct syntax and spacing. To prevent accidental overwrites, the argument `overwrite` is by default `FALSE`. Set it to `TRUE` to commit your changes and make them visible to the simulator, by overwriting the SWAT+ file on disk.

`rswat` relies on *attributes* to tie the return value of `rswat_open` back to its file of origin.

```{r show attributes}
attributes(print.prt.t1)
```

Be warned that some R functions drop attributes without warning (see `?Extract` for example). If you have issues with `rswat_write` after working on the output of `rswat_open`, check that you haven't accidentally discarded the `rswat` attributes.  
